Simple Wind Models

### 2010-07-06 Winds/papers/Theory/First/wind_simple.tex
blah
git-svn-id: https://svn.ucolick.org/xsvn/trunk/radtrans@5030 ac11df5e-a788-c23b-a8e1-c758c323924d
blah
git-svn-id: https://svn.ucolick.org/xsvn/trunk/radtrans@5031 ac11df5e-a788-c23b-a8e1-c758c323924d
blah
git-svn-id: https://svn.ucolick.org/xsvn/trunk/radtrans@5032 ac11df5e-a788-c23b-a8e1-c758c323924d

\newcommand{\taud}{$\tau_{\rm dust}$}
\newcommand{\mtaud}{\tau_{\rm dust}}
\newcommand{\maconfig}{2p$^6$3s}
\newcommand{\mbconfig}{2p$^6$3p}
\newcommand{\aconfig}{a~$^6$D$^0$}
\newcommand{\zconfig}{z~$^6$D$^0$}
\newcommand{\mvr}{v_{\rm r}}
\newcommand{\naid}{\ion{Na}{1}~$\lambda\lambda 5891, 5897$}
\newcommand{\mgiid}{\ion{Mg}{2}~$\lambda\lambda 2796, 2803$}
\newcommand{\mgiia}{\ion{Mg}{2}~$\lambda 2796$}
\newcommand{\mgiib}{\ion{Mg}{2}~$\lambda 2803$}
\newcommand{\feiia}{\ion{Fe}{2}~$\lambda 2586$}
\newcommand{\feiib}{\ion{Fe}{2}~$\lambda 2600$}
\newcommand{\feiic}{\ion{Fe}{2}$^* \; \lambda 2612$}
\newcommand{\feiid}{\ion{Fe}{2}~$\lambda\lambda 2586, 2600$}
\newcommand{\nmg}{$n_{\rm Mg^+}$}
\newcommand{\mnmg}{n_{\rm Mg^+}}
\newcommand{\nfe}{$n_{\rm Fe^+}$}
\newcommand{\mnfe}{n_{\rm Fe^+}}
\def\hub{h_{72}^{-1}}
\def\umfp{{\hub \, \rm Mpc}}
\def\mzq{z_q}
\def\zabs{$z_{\rm abs}$}
\def\mzabs{z_{\rm abs}}
\def\intl{\int\limits}
\def\cmma{\;\;\; ,}
\def\perd{\;\;\; .}
\def\ltk{\left [ \,}
\def\ltp{\left ( \,}
\def\ltb{\left \{ \,}
\def\rtk{\, \right  ] }
\def\rtp{\, \right  ) }
\def\rtb{\, \right \} }
\def\sci#1{{\; \times \; 10^{#1}}}
\def \rAA {\rm \AA}
\def \zem {$z_{\rm em}$}
\def \mzem {z_{\rm em}}
\def\smm{\sum\limits}
\def \cmm  {cm$^{-2}$}
\def \cmmm {cm$^{-3}$}
\def \kms  {km~s$^{-1}$}
\def \mkms  {{\rm km~s^{-1}}}
\def \lyaf {Ly$\alpha$ forest}
\def \Lya  {Ly$\alpha$}
\def \lya  {Ly$\alpha$}
\def \mlya  {Ly\alpha}
\def \Lyb  {Ly$\beta$}
\def \lyb  {Ly$\beta$}
\def \lyg  {Ly$\gamma$}
\def \ly5  {Ly-5}
\def \ly6  {Ly-6}
\def \ly7  {Ly-7}
\def \nhi  {$N_{\rm HI}$}
\def \mnhi  {N_{\rm HI}}
\def \lnhi {$\log N_{HI}$}
\def \mlnhi {\log N_{HI}}
\def \etal {\textit{et al.}}
\def \lyaf {Lyman--$\alpha$ forest}
\def \mnmin {\mnhi^{\rm min}}
\def \nmin {$\mnhi^{\rm min}$}
\def \O {${\mathcal O}(N,X)$}
\newcommand{\cm}[1]{\, {\rm cm^{#1}}}
\def \snrlim {SNR$_{lim}$}

@@

Our 3D Monte Carlo algorithm tracks the spatial emissivity of the source+wind
complex as a function of frequency.  From this output, we
constructed surface-brightness maps producing a
dataset analagous to intergral-field-unit (IFU) observations.  In
Figure~\ref{fig:fiducial_ifu_mgii}, we present the output for our
fiducial wind model at several velocities relative to the \mgiia\
transition. At $\mvr \le -200 \mkms$, where the wind has an optical
depth $\tau_{2796} < 1$ (Figure~\ref{fig:fig_fiducial_nvt}),
the source contributes significantly to the observed flux.  
At $\mvr=-100 \mkms$, however, the
wind absorbs all photons from the source and the observed emission is
entirely from photons scattered by the wind.  This scattered emission
actually exceeds the source+wind emission at 
$\mvr = -200\mkms$ such
that the absorption profile reaches peak depth at a velocity that is
negatively offset from
the peak in optical depth.  The net result of the scattered photons is
a weaker \ion{Mg}{2} profile that is offset fromt he true optical
depth profile.  Clearly, this will complicate efforts to estimate the
speed, covering fraction, and total column density of the wind.  At $v
= 0 \, \mkms$, the wind and source have comparable total flux and the
latter dominates at higher velocities.  
At $v_{\rm r} \ge 0
\mkms$,  both the source and wind contribute to the observed emission.

@@

\subsection{Dust}

@@

@@

\input{Tables/tab_fiducial.tex}
\input{Tables/tab_fiducial_ew.tex}

@@

>>>>

\newcommand{\taud}{$\tau_{\rm dust}$}
\newcommand{\mtaud}{\tau_{\rm dust}}
\newcommand{\maconfig}{2p$^6$3s}
\newcommand{\mbconfig}{2p$^6$3p}
\newcommand{\aconfig}{a~$^6$D$^0$}
\newcommand{\zconfig}{z~$^6$D$^0$}
\newcommand{\mvr}{v_{\rm r}}
\newcommand{\naid}{\ion{Na}{1}~$\lambda\lambda 5891, 5897$}
\newcommand{\mgiid}{\ion{Mg}{2}~$\lambda\lambda 2796, 2803$}
\newcommand{\mgiia}{\ion{Mg}{2}~$\lambda 2796$}
\newcommand{\mgiib}{\ion{Mg}{2}~$\lambda 2803$}
\newcommand{\feiia}{\ion{Fe}{2}~$\lambda 2586$}
\newcommand{\feiib}{\ion{Fe}{2}~$\lambda 2600$}
\newcommand{\feiic}{\ion{Fe}{2}$^* \; \lambda 2612$}
\newcommand{\feiie}{\ion{Fe}{2}$^* \; \lambda 2626$}
\newcommand{\feiid}{\ion{Fe}{2}~$\lambda\lambda 2586, 2600$}
\newcommand{\nmg}{$n_{\rm Mg^+}$}
\newcommand{\mnmg}{n_{\rm Mg^+}}
\newcommand{\nfe}{$n_{\rm Fe^+}$}
\newcommand{\mnfe}{n_{\rm Fe^+}}
\def\hub{h_{72}^{-1}}
\def\umfp{{\hub \, \rm Mpc}}
\def\mzq{z_q}
\def\zabs{$z_{\rm abs}$}
\def\mzabs{z_{\rm abs}}
\def\intl{\int\limits}
\def\cmma{\;\;\; ,}
\def\perd{\;\;\; .}
\def\ltk{\left [ \,}
\def\ltp{\left ( \,}
\def\ltb{\left \{ \,}
\def\rtk{\, \right  ] }
\def\rtp{\, \right  ) }
\def\rtb{\, \right \} }
\def\sci#1{{\; \times \; 10^{#1}}}
\def \rAA {\rm \AA}
\def \zem {$z_{\rm em}$}
\def \mzem {z_{\rm em}}
\def\smm{\sum\limits}
\def \cmm  {cm$^{-2}$}
\def \cmmm {cm$^{-3}$}
\def \kms  {km~s$^{-1}$}
\def \mkms  {{\rm km~s^{-1}}}
\def \lyaf {Ly$\alpha$ forest}
\def \Lya  {Ly$\alpha$}
\def \lya  {Ly$\alpha$}
\def \mlya  {Ly\alpha}
\def \Lyb  {Ly$\beta$}
\def \lyb  {Ly$\beta$}
\def \lyg  {Ly$\gamma$}
\def \ly5  {Ly-5}
\def \ly6  {Ly-6}
\def \ly7  {Ly-7}
\def \nhi  {$N_{\rm HI}$}
\def \mnhi  {N_{\rm HI}}
\def \lnhi {$\log N_{HI}$}
\def \mlnhi {\log N_{HI}}
\def \etal {\textit{et al.}}
\def \lyaf {Lyman--$\alpha$ forest}
\def \mnmin {\mnhi^{\rm min}}
\def \nmin {$\mnhi^{\rm min}$}
\def \O {${\mathcal O}(N,X)$}
\newcommand{\cm}[1]{\, {\rm cm^{#1}}}
\def \snrlim {SNR$_{lim}$}

@@

Our 3D Monte Carlo algorithm tracks the spatial emissivity of the source+wind
complex as a function of frequency.  From this output, we
constructed surface-brightness maps producing a
dataset analagous to intergral-field-unit (IFU) observations.  In
Figure~\ref{fig:fiducial_ifu_mgii}, we present the output for our
fiducial wind model at several velocities relative to the \mgiia\
transition. At $v \le -200 \mkms$, where the wind has an optical
depth $\tau_{2796} < 1$ (Figure~\ref{fig:fig_fiducial_nvt}),
the source contributes significantly to the observed flux.  
At $v=-100 \mkms$, however, the
wind absorbs all photons from the source and the observed emission is
entirely from photons scattered by the wind.  This scattered emission
actually exceeds the source+wind emission at 
$v = -200\mkms$ such
that the absorption profile reaches peak depth at a velocity that is
negatively offset from
the peak in optical depth.  The net result of the scattered photons is
a weaker \ion{Mg}{2} profile that is offset fromt he true optical
depth profile.  Clearly, this will complicate efforts to estimate the
speed, covering fraction, and total column density of the wind.  At $v
= 0 \, \mkms$, the wind and source have comparable total flux and the
latter dominates at higher velocities.  
At $v_{\rm r} \ge 0
\mkms$,  both the source and wind contribute to the observed emission.

@@

\subsection{Dust}
\label{sec:dust}

@@

\subsection{ISM}
\label{sec:ISM}

As designed, the fiducial mdoel does not include gas associated with
the interstellar medium of the galaxy, i.e.\ material at $r \sim
0$\,kpc with $v_{\rm r} \approx 0 \mkms$.  This placed the focus on
properties of the wind.  In part, this decision was also motivated by
the general absence of significant absorption at $v \approx 0 \mkms$
in galaxies that exhibit outflows
\citep[e.g.][]{wcp+09,steidel+10,rubin+10}.
[Create an ISM grid for Dan]
On the other hand, the stars and \ion{H}{2} regions that comprise our
background sources are very likely embedded within and fueled by gas
of the galactic ISM.

Consider a modification to the fiducial model that adopts the same
wind but also includes an ISM component with $<\mvr> = 0 \mkms$.  
Specifically, the ISM component has $n_{\rm H} = 1 \cm{-3}$ for
$r_{\rm ISM} \le r < r_{\rm inner}$, an average velocity of $\mvr = 0
\mkms$, but a large turbulent velocity $b_{\rm ISM} = 40 \mkms$.
Figure~\ref{fig:ISM} summarizes the properties of this model.
[Comment?]

In Figure~\ref{fig:ISM_spec} the solid curves show the \ion{Mg}{2} and
\ion{Fe}{2} profiles for the ISM+wind (black) and fiducial (red)
models. In comparison, the
dotted curve shows the absorption profile for the ISM+wind in the (unphysical) case
where no absorbed photons are re-emitted.   Focus first on the
\ion{Mg}{2} doublet.  As expected, the dotted curve shows strong
absorption at $v \approx 0 \mkms$ and blueward.  The full models,
in contrast, show positive emission at $v \approx 0 \mkms$.  In
fact, the ISM+wind model is nearly identical to the fiducial model
with the only quantitiative difference being that the velocity centroid of
the emission lines are shifted by $\approx +100 \mkms$.

There are, however, several qualitiative differences between the
ISM+wind and fiducial models for the \ion{Fe}{2} transitions.  First,
the \feiia\ transition shows much stronger absorption at $v \approx 0$
to $-100 \mkms$.  The difference from the \ion{Mg}{2} doublet is that
this absorption is not filled in by scattered photons because the
majority of photons absorbed at the \feiia\ are re-emitted as
\ion{Fe}{2}~$^* \lambda\lambda 2612, 2632$ photons.  In fact, the
\feiia\ profile with scattering nearly matches that withou re-emission
(compare the solid and dotted lines) and this transition provides a
very good description of the ISM+wind opacity.  The other key
difference is the much higher strength ($3-10\times$) of emission from the
transitions to the excited states of \aconfig.   There are two reasons
for this higher emission: (1) the absorption is greater in the \feiid\
resonance lines; (2) the high opacity of the ISM component leads to
the increased conversion of resonance photons with $v \approx 0\mkms$
into the non-resonance lines.  This is especially notable for the
\feiib\ transition whose `partner' shows an equivalent width nearly
$10\times$ stronger than for the fiducial model.  The relative
strengths of the \feiib\ and \feiie\ lines therefore provide a direct
diagnostic on the degree to which the resonance lines are trapped.
[Comment:  this is a product of $\tau_{2600}$ and the velocity
gradient]

\subsection{The Lyman Break Galaxy Model}
\label{sec:lbg}

The Lyman break galaxies, UV color-selected galaxies at $z \sim 3$,
are observerd to exhibit cool gas outflows in \ion{Si}{2}, \ion{C}{2},
etc.\ transitions with maximum speeds of nearly 1000\kms\
\citep[e.g.][]{lowenthal,pettini,steidel+10}.
Recently, \cite{steidel+10} modeled the average absorption observed
toward a set of XX LBGs and the average absorption in cool gas
observed transverse to these galaxies.  Using our methodology, we have
explored further the wind that they inferred from the observations.

\cite{steidel+10} introduced a wind model parameterized by two
quantities: a radial velocity low $\mvr(r)$ and the covering fraction of
optically thick gas $f_c(r)$.  [Explain the origin of these better?]
For the velocity law, the authors introduced the following functional
form

\begin{equation}
\mvr = \ltp \frac{A_{\rm LBG}}{1-\alpha} \rtp^{1/2} \ltk r_{\rm
  inner}^{1-\alpha} - r^{1-\alpha} \rtk^{1/2}
\label{eqn:LBG_vlaw}
\end{equation}
with $A_{\rm LBG}$ a constant that sends the terminal speed,
$r_{\inner}$ is the inner radius of the wind (taken to be 1\,kpc) and
$\alpha$ describes how steeply the velocity curve rises.  Their
analysis of the LBG absorption profiles required
a very steeply rising curve with $\alpha \approx 1.3$.
The resulting curve is shown as a dotted line in the top panel of
Figure~\ref{fig:LBG_Sobolev}.

The covering fraction of the cold gas, meanwhile, was assumed to have
the functional form

\begin{equation}
f_c(r) = f_{c,max} \ltp \frac{r}{r_{\rm inner}} \rtp^{-\gamma}
\label{eqn:covering}
\end{equation}
with $\gamma \approx 0.5$ and $f_{c,max}$ the maximum covering
fraction.  From this expression and the velocity law, one can recover
an absorption profile $I(v) = 1 - f_c(r[v])$.  Inverting the velocity
law, we derive

\begin{equation}
I_{\rm LBG}(v) = 1 - f_{c,max} \ltk r_{\rm inner}^{1-\alpha} - \ltp
\frac{1-\alpha}{A_{\rm LBG}} \rtp v^2 \rtk^{\frac{\gamma}{\alpha-1}}
\perd
\label{eqn:LBG_I}
\end{equation}
The resulting profile for $f_{c,max} = 0.6$, $\gamma=0.5$,
$\alpha=1.3$, and $A_{\rm LBG} = $ is displayed in the lower panel of
Figure~\ref{fig:LBG_Sobolev}.  
In the following, we consider two methods to analyze the LBG wind.
Both approaches assume isotropy and adopt the velocity law given by
Equation~\ref{eqn:LBG_vlaw}.  In one model, we treat the cool gas as a
diffuse medium with a radial density profile.  We then apply 
the methodology used for the other wind models.  In the other model,
we modify our algorithms to more precisely mimic the model suggested
by \cite{steidel+10}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Sobolev}
\label{sec:Sobolev}

As demonstrated in Figure~\ref{fig:LBG_Sobolev}, the wind velocity for
the LBG model rises very steeply with increasing radius before
flattening at large radii.  Under these conditions, the Sobolev
approximation can be applied to derive the density profile provided
one has a description of the optical depth profile $\tau(v)$.
For the latter, we take

\begin{equation}
\tau_{\rm LBG}(v) = -\ln \ltk I_{\rm LBG}(v) \rtk
\label{eqn:tauLBG}
\end{equation}
and one can express the optical depth in terms of the radius using
Equation~\ref{eqn:LBG_vlaw}.  Following \cite{sobolev60}, we have

\begin{equation}
n_{\rm LBG}(r) = \frac{\tau_{\rm LBG}(r) \; d\mvr/dr}{\kappa_\ell
  \lambda_0}
\label{eqn:Sobolev}
\end{equation}
with $\lambda_0$ the rest wavelength and $\kappa_\ell \equiv f\pi^2
e^2/m_e c$.  The solid curve in the top panel of Figure~\ref{fig:LBG_Sobolev} shows the
resultant density profile for Hydrogen.  [Probably show $n_{\rm
  Mg^+}$]
[This is a relatively extreme density profile.  From the inner radius
of the wind (1\,kpc) to 2\,kpc, the density drops by over 2 orders of
magnitude including approximately an order of magnitude over the first
10\,pc.  Beyond 2\,kpc, the desnity drops even more rapidly, falling
orders of magnitude from 2 to 100\,kpc.]
It was derived by assuming
that \mgiia\ transition has the optical depth profile given in
Equation~\ref{eqn:tauLBG} (the black curve in the lower panel of
Figure~\ref{fig:LBG_Sobolev})  and then scaling by $10^{5.77}$ which 
corresponds to Mg depleted by a factor of 10 and a 1/2 solar
metallicity.  [Give more detail?  Probably]
We then verified that this density profile reproduced the correct
absorption profile by discretizing the gas into a series of tiny
`clouds' and calculating the integrated absorption profile.  This is
overplotted as a dotted red curve on the lower which is an excellent
match to the desired profile (black curve).
For Fe$^+$, we take $n_{\rm Fe^+} = n_{\rm Mg^+}/2$ and the strength
of all other transitions scale as $f\lambda$, i.e.\ the 
\mgiib\ profile will be half as deep as that shown in
Figure~\ref{fig:LBG_Sobolev}.

We then generated \ion{Mg}{2} and \ion{Fe}{2} profiles for this wind
using the 1D algorithm with no dust extinction; these are shown with
the red curves in Figure~\ref{fig:LBG_spec}.   For this analysis, one
should focus on the \mgiia\ transition.  The dotted line in the Figure
shows the intensity profile for the model when one ignores re-emission
of absorbed photons.  By construction, it follows\footnote{Note that
  one should not make this comparison for the other transitions in
  this Sobolev model because these are scaled down by $f\lambda$ and
  the reduced Fe$^+$ abundance.} the profile
described in Equation~\ref{eqn:LBG_I}.   The full model, however,
shows much weaker absorption especially at $v = 0$ to $-300 \mkms$.
In this respect, the LBG-Sobolev model is an
inaccurate description of the observations.  In addition, the model
predicts significant emission in the \ion{Mg}{2} lines and several of
the \ion{Fe}{2}$^*$ transitions.   Emission associated with cool gas
has been observed for non-resonant \ion{Si}{2}
transitions in LBGs \citep{cb58,shapley}.
The \ion{Fe}{2} and \ion{Mg}{2} transitions
  modeled here lie in the near-IR and have not yet been [explored].
  [X-Shooter must help here]  On the other hand, there have been no 
reported detections of significant line-emission related to resonance
transitions in LBGs, only $z \le 1$ star-forming galaxies
\citep{wcp+09,rubin09}.  We discuss these points further in the
following section. [Do we?]

[Does CIV scattered photons bugger up the P-Cygni analysis of Pettini
et al. ?]

[Does most of the scattering occur within the inner few kpc? Explore
this with outflow.cc]

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Covering Fraction}
\label{sec:Covering}

In the previous sub-section, we described a Sobolev solution that
reproduces the average absorption profile of LBGs in cool gas
transitions.  We found that a full analysis of this wind model leads
to absorption profiles that are qualitatively different from the ones
observed.  This is due to the re-emission and scattering of photons
within the wind itself.  We also noted that this model differed from
the one proposed by S10.  Their model [argued for] discrete, optically
thick clouds with a non-unity covering fraction (described by
Equation~\ref{eqn:covering}).   In contrast, the Sobolev model assumes
a diffuse medium with a declining density profile but a unit covering
fraction.

To more properly model the LBG wind described in S10, we 
performed the following Monte Carlo calculation.  First, we propogated
a photon from the source until its velocity relative to line-center
resonates with the wind (it escapes if this never occurs).
The photon then has a probability $P = f_c(r)$ of scattering.  If it
scatters, we continue to track the photon until it comes into
resonance again or escapes the system.  In this model, all of the
resonance transitions are assumed to have identical optical depth in
the absence of re-emission (the dotted line in
Figure~\ref{fig:LBG_spec}).

@@

\input{Tables/tab_fiducial.tex}
\input{Tables/tab_fiducial_ew.tex}
\input{Tables/tab_meas_models.tex}
\input{Tables/tab_meas_plaws.tex}

@@

\begin{figure}
\epsscale{0.8}
\plotone{Figures/fig_ism_diagn.ps}
\caption{
Density (solid; black), radial velocity (dotted; blue), and
\mgiia\ optical depth profiles (dashed; red) for the ISM+wind
model.  The wind is identical to the fiducial model
(Figure~\ref{fig:fiducial_nvt}) but we also include an ISM component
with $n_{\rm H} = 1 \cm{-3}$ that has an average velocity of $\mvr = 0
\mkms$ and a large turbulent velocity $b_{\rm ISM} = 40 \mkms$.
Note that the density and velocity curves have been scaled for plotting
convenience.  
}
\label{fig:ISM}
\end{figure}

\begin{figure}
\epsscale{0.8}
\plotone{Figures/fig_ism_spec.ps}
\caption{
Profiles of the \ion{Fe}{2} and \ion{Mg}{2} profiles for the ISM+wind
model (black) compared against the fiducial wind model (red). 
The dotted line traces the predicted absorption profile in the absence
of any re-emission of absorbed photons.
Regarding the \ion{Mg}{2}~doublet, the primary difference between the
ISM+wind and the fiducial models is the shift of $\approx 100 \mkms$
in the emission lines from $\mvr \approx 0 \mkms$ for the fiducial
model to $\mvr \approx +100\mkms$ for the ISM+wind model. 
[Discuss FeII]
}
\label{fig:ISM_spec}
\end{figure}

\begin{figure}
\epsscale{0.8}
\plotone{Figures/fig_lbg_sobolev.ps}
\caption{
{\it Upper:} The dotted curve shows the velocity law for the LBG model
of \cite{steidel+10}.  Note how rapidly the velocity rises from $r =
1$ to 2\,kpc.  The solid curve shows the Sobolev solution required to
match the average absorption profile of LBG galaxies (lower panel).
The density drops off as $(r-1)^{-0.6}$ 
{\it Lower:} Absorption profiles for the LBG model as defined by
\cite{steidel+10} (black solid curve) and for our Sobolev approximation of
the model (red dotted curve).  
}
\label{fig:LBG_Sobolev}
\end{figure}

\begin{figure}
\epsscale{0.8}
\plotone{Figures/fig_lbg_spec.ps}
\caption{
Profiles of the \ion{Fe}{2} and \ion{Mg}{2} profiles for the fiducial
}
\label{fig:LBG_spec}
\end{figure}
